Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-01 LaTeX/Turnover.tex

If there is genetic variation maintained by sexually-antagonistic selection, new sex-determining systems can evolve. 
In particular, transitions to new sex-determining systems can occur when new sex-determining regions are more closely linked to a sexually-antagonistic locus.
Here, we show that genetic variation at loci that experience haploid selection can also generate selection in favour of new sex-determining systems. 
Closer linkage can again favour new sex-determining alleles. 
However, with haploid selection, heterogametic transitions (XY to ZW or ZW to XY) can also occur when the new sex-determining region is less closely linked to the locus under selection.

Neo-W (neo-Y) alleles invade when their fitness in females (males) is greater than the mean fitness of females (males) under the ancestral sex determination system.
With sexually antagonistic selection (between diploid sexes) only, linkage between a selected locus and the sex-determining region strengthens associations between male beneficial alleles and the male-determining allele (Y or Z) and between female beneficial alleles and the female-determining allele (X or W). 
Thus, the mean fitness of both males and females increases with closer linkage to the sex-determining region. 
Therefore, new sex-determining alleles only invade if they are more closely linked than the ancestral sex-determining region. 
However, if there is haploid selection on loci linked to an XY (ZW) sex-determining region, there are equilibria for which the mean fitness of females (males) or males is lower than it would be without sex-linkage, allowing unlinked neo-W (neo-Y) alleles to invade, see figure \ref{fig:Combination_MeanFit}.

Linkage between a locus that experiences haploid selection and a sex-determining region causes zygotic sex ratios to become biased \textcolor{red}{(REFERENCES)}, which we might then expect to exert sex ratio selection upon new modifiers.
Fisherian sex ratio selection follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male, and half from a female. 
Thus, if the population sex ratio is biased towards females, the average per-individual contribution of genetic material to the next generation from males is greater than the contribution from females (and vice versa for male-biased sex ratios). 
Therefore, a mutant that increases investment in males will spread via the higher per-individual contributions made by males. 
That is, under Fisherian sex ratio selection, the success of a mutant relative to the non-mutant depends, in equal parts, on the contributions made by males and females to the next generation. 
An implicit assumption of Fisherian sex ratio selection is that the mutant allele is autosomal and has the same inheritance pattern as the non-mutant allele. 
The mutations we consider here, neo-sex-determining alleles, break this assumption. 
For example, the success of neo-Y mutations depends only on the number of alleles contributed by males (equation \ref{eq:charpoly_neoY} and Table \ref{tab:haplotype_growth}). 
Even mutants that are equally likely to be found in males or females, such as an environmental sex determination mutation (equation \ref{eq:lambda_ESD}), are not strictly autosomal if they determine sex. 
Thus, despite the fact that sex ratio biases caused by haploid competition or meiotic drive have been shown to exert selection on various autosomal modifiers \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va}, we do not find evidence of Fisherian sex ratio selection acting upon neo-sex-determination systems (e.g., see Figure \ref{fig:Combination_Turnover}).

>>>>

One might expect Fisherian sex ratio selection to influence the spread of new sex-determining systems because linkage between haploid selected loci and sex-determining regions cause biased zygotic sex ratios \citep{Hamilton:1967ts,Burt:2006,Field:2012fd,Field:2013cc}. 
Fisherian sex ratio selection follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male, and half from a female. 
Thus, if the population sex ratio is biased towards females, the average per-individual contribution of genetic material to the next generation from males is greater than the contribution from females (and vice versa for male-biased sex ratios). 
Therefore, a mutant that increases investment in males will spread via the higher per-individual contributions made by males. 
That is, under Fisherian sex ratio selection, the success of a mutant relative to the non-mutant depends, in equal parts, on the contributions made by males and females to the next generation. 
An implicit assumption of Fisherian sex ratio selection is that the mutant allele is autosomal and has the same inheritance pattern as the non-mutant allele. 
The mutations we consider here, neo-sex-determining alleles, break this assumption. 
For example, the success of neo-Y mutations depends only on the number of alleles contributed by males (equation \ref{eq:charpoly_neoY} and Table \ref{tab:haplotype_growth}). 
Even mutants that are equally likely to be found in males or females, such as an environmental sex determination mutation (equation \ref{eq:lambda_ESD}), are not strictly autosomal if they determine sex. 
Thus, despite the fact that sex ratio biases caused by haploid competition or meiotic drive have been shown to exert selection on various autosomal modifiers \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va}, we do not find evidence of Fisherian sex ratio selection acting upon neo-sex-determination systems (e.g., see Figure \ref{fig:Combination_Turnover} and \citealt{Ubeda:2015fx}, in which a neo-Y invades despite biasing sex ratios).

It has previously been demonstrated that new sex-determining systems can evolve if there is genetic variation maintained by sexually-antagonistic selection \citep{vanDoorn:2007eu,vanDoorn:2010hu}. 
In particular, transitions to new sex-determining systems can occur when new sex-determining regions are more closely linked to a sexually-antagonistic locus.
Our results show that genetic variation at loci that experience haploid selection can also generate selection in favour of new sex-determining systems. 
New sex-determining allleles are again favoured if they are more closely linked with a locus under haploid selection.
However, with haploid selection, heterogametic transitions (XY to ZW or ZW to XY) can also occur when the new sex-determining region is less closely linked to the locus under selection.

Neo-W (neo-Y) alleles invade when their fitness in females (males) is greater than the mean fitness of females (males) under the ancestral sex determination system.
With sexually antagonistic selection (between diploid sexes) only, linkage between a selected locus and the sex-determining region strengthens associations between male beneficial alleles and the male-determining allele (Y or Z) and between female beneficial alleles and the female-determining allele (X or W). 
Thus, the mean fitness of both males and females increases with closer linkage to the sex-determining region. 
Therefore, new sex-determining alleles only invade if they are more closely linked than the ancestral sex-determining region. 
However, if there is haploid selection on loci linked to an XY (ZW) sex-determining region, polymorphisms can be maintained at which the mean fitness of females (males) or males is lower than it would be without sex-linkage, allowing unlinked neo-W (neo-Y) alleles to invade, see figure \ref{fig:Combination_MeanFit}.
